---
output: 
  github_document:
    html_preview: false
editor_options: 
  chunk_output_type: console
---

This repository provides a standardized archive of the US Department of Agriculture Farm Service Agency (FSA) [Farm Payment Files](https://www.fsa.usda.gov/tools/informational/freedom-information-act-foia/electronic-reading-room/frequently-requested/payment-files). These files contain detailed records of payments made to agricultural program participants, originally released as Microsoft Excel files.

The included R script automates the discovery, download, and conversion of these files into a partitioned, analysis-ready format.

---

## 📁 Repository Contents

- `fsa-payment-files.R`: Main R script to download, extract, clean, and write the payment files to a partitioned Parquet format.
- `README.Rmd`: This file. Contains background information, usage instructions, and archive structure.
- `data-raw/`: Folder used for downloading and unpacking Excel and CSV files.
- `fsa-payment-files/`: Final output directory containing the processed data as a Parquet dataset, partitioned by state and year. When published to S3, this prefix also carries a `_manifest.txt` listing the HTTPS URL of every Parquet file (the leading underscore keeps Arrow's dataset discovery from reading it as data).

---

## 🔁 Processing Workflow

The `fsa-payment-files.R` script performs the following steps:

1. **Discovery**: Scrapes the FSA payment files website to find downloadable Excel files (2004–2025).
2. **Download**: Saves original files to `data-raw/`.
3. **Extraction**: Parses each Excel file, handling annual variation in schema and formatting.
4. **Standardization**: Renames fields, standardizes formats (e.g., FIPS codes, program names), and removes duplicates.
5. **Archiving**: Writes the full dataset to the `fsa-payment-files/` directory as partitioned Parquet files (`State FSA Name=` and `Accounting Program Year=` directories).
6. **Upload** *(optional)*: Syncs the Parquet archive to a public S3 bucket with the AWS CLI (`aws s3 sync`), verifies the remote listing matches the local build, and publishes a `_manifest.txt` of HTTPS URLs for access via the CloudFront portal.

---

## ☁️ Public Access via S3

The full archive is hosted in a public Amazon S3 bucket:

```
s3://sustainable-fsa/fsa-payment-files/
```

You can access the data directly using:

### AWS CLI
```bash
aws s3 ls s3://sustainable-fsa/fsa-payment-files/ --no-sign-request
```

### In R with `arrow` (recommended)
```{r arrow-s3}
library(arrow)

bucket <- s3_bucket("sustainable-fsa", anonymous = TRUE, region = "us-west-2")

dataset <- open_dataset(bucket$path("fsa-payment-files"))

dataset
```

Queries that filter on the partition columns (`State FSA Name`, `Accounting Program Year`) only download the matching Parquet files. Arrow decodes the partition values for you, so filters use plain names — e.g., `dplyr::filter(\`State FSA Name\` == "New Mexico")`.

---

## 🌐 Public Access via HTTPS (CloudFront)

The same archive is served over plain HTTPS at:

```
https://data.sustainable-fsa.com/fsa-payment-files/
```

HTTPS has no directory listing, so a manifest of every file URL is published at [`_manifest.txt`](https://data.sustainable-fsa.com/fsa-payment-files/_manifest.txt) (regenerated on every data update). [DuckDB](https://duckdb.org)'s `httpfs` extension can query these URLs directly — and filters on the hive partition columns are applied to the URL paths *before* any file is fetched, so you keep the benefits of the partitioning without S3 access:

```{r duckdb-cloudfront}
library(DBI)

manifest <- readLines("https://data.sustainable-fsa.com/fsa-payment-files/_manifest.txt")

con <- dbConnect(duckdb::duckdb())
invisible(dbExecute(con, "INSTALL httpfs; LOAD httpfs;"))

files_sql <- paste0("[", paste0("'", manifest, "'", collapse = ",\n"), "]")

lfp_mt <- dbGetQuery(con, glue::glue('
  SELECT "Accounting%20Program%20Year" AS "Accounting Program Year",
         SUM("Disbursement Amount") AS "Disbursement Amount"
  FROM read_parquet({files_sql}, hive_partitioning = true)
  WHERE "State%20FSA%20Name" = \'Montana\'
  GROUP BY 1
  ORDER BY 1
'))

dbDisconnect(con, shutdown = TRUE)

tail(lfp_mt)
```

Notes for DuckDB users:

- Because these URLs are percent-encoded, the hive column names surface encoded — `"State%20FSA%20Name"` and `"Accounting%20Program%20Year"` (quote them with *double* quotes in SQL) — and multi-word partition values keep one layer of encoding, so filters need encoded literals (e.g., `WHERE "State%20FSA%20Name" = 'New%20Mexico'`). Use DuckDB's `url_decode()` to clean values for display.
- `"Accounting%20Program%20Year"` autocasts to an integer. A few partitions hold rows with malformed years in the source files: `__HIVE_DEFAULT_PARTITION__` (surfaces as `NULL`) and `0` — filter them out if they matter for your analysis.
- For simple cases you can skip the SQL-side filter entirely and subset the manifest in R first, e.g. `manifest[grepl("Name=Montana/", manifest, fixed = TRUE)]`.
- DuckDB can also query the S3 bucket directly — no manifest needed, and here the column names and values come through clean (`"State FSA Name"`, `'New Mexico'`):

```r
con <- dbConnect(duckdb::duckdb())
dbExecute(con, "INSTALL httpfs; LOAD httpfs;")
dbExecute(con, "CREATE SECRET (TYPE s3, PROVIDER config, REGION 'us-west-2');")
dbGetQuery(con, '
  SELECT "County FSA Name", SUM("Disbursement Amount") AS total
  FROM read_parquet(\'s3://sustainable-fsa/fsa-payment-files/*/*/*.parquet\',
                    hive_partitioning = true)
  WHERE "State FSA Name" = \'Montana\'
  GROUP BY 1 ORDER BY total DESC LIMIT 10')
```

---

## 🧭 Notes

- File structure varies by year, and the script includes heuristics to detect column shifts and missing headers.
- Partitioning by state and year facilitates fast querying and cloud-native workflows.
- The processing script can be rerun to include new years as they are released.

---

## 📅 Update Schedule

This dataset is refreshed annually after the USDA FSA releases new payment files, typically in the spring. Additional years or corrections may be processed as available.

---

## 🔗 Related Resources

- [USDA FSA FOIA Payment Page](https://www.fsa.usda.gov/news-room/efoia/electronic-reading-room/frequently-requested-information/payment-files-information/index)
- [Arrow and Parquet in R](https://arrow.apache.org/docs/r/)

---

## 📍 Quick Start: Visualize data in the FSA Farm Payment Files in R

This snippet shows how to load data from the Farm Payment Files archive and create a simple map using `sf` and `ggplot2`.

```{r example, message=FALSE, fig.align = 'center', dpi=300, fig.path = './'}
# Load required libraries
library(arrow)
library(sf)
library(ggplot2) # For plotting
library(tigris)  # For state boundaries
library(rmapshaper) # For innerlines function

# Example accessing payment files on S3
# A map of 2025 LFP Payments by county

lfp_payments <-
  # To run this from anywhere, open the dataset from S3 instead:
  # arrow::s3_bucket("sustainable-fsa", anonymous = TRUE,
  #                  region = "us-west-2")$path("fsa-payment-files") |>
  "fsa-payment-files" |>
  arrow::open_dataset() |>
  dplyr::filter(`Accounting Program Description` %in% 
                  c(
                    "LIVESTOCK FORAGE PROGRAM",
                    "LIVESTOCK FORAGE DISASTER PROGRAM",
                    "LIVESTOCK FORAGE DISASTER PROGRAM (COF)"
                  )) |>
  dplyr::group_by(`FSA Code`) |>
  dplyr::summarise(
    `Disbursement Amount` = sum(`Disbursement Amount`, na.rm = TRUE)
  ) |>
  dplyr::collect()

## Download from the FSA_Counties_dd17 archive
counties <- 
  sf::read_sf("https://data.sustainable-fsa.com/fsa-counties-dd17/fsa-counties-dd17.topojson",
              layer = "counties") |>
  sf::st_set_crs("EPSG:4326") |>
  sf::st_transform("EPSG:5070")

lfp_payments_counties <-
  lfp_payments |>
  dplyr::select(id = `FSA Code`,
                `Disbursement Amount`) |>
  dplyr::right_join(counties) |>
  sf::st_as_sf() |>
  dplyr::mutate(
    `Disbursement Amount` = 
      tidyr::replace_na(`Disbursement Amount`, 0)
  )


# Plot the map
ggplot(counties) +
  geom_sf(data = sf::st_union(counties),
          fill = "grey80",
          color = NA) +
  geom_sf(data = lfp_payments_counties,
          aes(fill = `Disbursement Amount`), 
          color = NA) +
  geom_sf(data = rmapshaper::ms_innerlines(counties),
          fill = NA,
          color = "white",
          linewidth = 0.1) +
  geom_sf(data = counties |>
            dplyr::group_by(state) |>
            dplyr::summarise() |>
            rmapshaper::ms_innerlines(),
          fill = NA,
          color = "white",
          linewidth = 0.2) +
  scale_fill_viridis_c(#palette = "OrRd",
    option = "turbo",
    direction = 1,
    na.value = "grey80",
    limits = c(0,100000000),
    breaks = c(0,10000,100000,1000000,10000000, 100000000),
    trans = scales::pseudo_log_trans(sigma = 1000),
    labels=scales::label_currency(scale_cut = scales::cut_short_scale()),
    name = 
      paste0(
        "Livestock Forage Program Payments, 2008–2025\nTotal: ",  
        scales:::dollar(sum(lfp_payments_counties$`Disbursement Amount`, na.rm = TRUE)),
        " as of ", format(lubridate::today(), "%B %d, %Y")
      ),
    guide = guide_colorbar(title.position = "top") ) +
  theme_void(base_size = 24) +
  theme(legend.position = "bottom",
        legend.justification = "center",
        legend.key.width = unit(0.15, "npc"),
        legend.title = element_text(size = 16, hjust = 0.5),
        legend.text = element_text(size = 14, hjust = 0.5),
        strip.text.x = element_text(margin = margin(b = 5)),
        strip.text.y = element_text(margin = margin(r = 5)))
```

---

## 🧭 About FSA County Codes

The USDA FSA uses custom county definitions that differ from standard ANSI/FIPS codes used by the U.S. Census. To align the Farm Payment Files with geographic boundaries, use the FSA-specific geospatial dataset archived in the companion repository:

🔗 [**sustainable-fsa/fsa-counties-dd17**](https://sustainable-fsa.com/fsa-counties-dd17/)

FSA county codes are documented in [FSA Handbook 1-CM, Exhibit 101](https://www.fsa.usda.gov/Internet/FSA_File/1-cm_r03_a80.pdf).

---

## 📝 Citation

If you use this data in published work, please cite:

> USDA Farm Service Agency. *Farm Payment Files, 2004–2025*. Curated and archived by R. Kyle Bocinsky, Montana Climate Office, University of Montana. Sustainable FSA project. Accessed YYYY-MM-DD. <https://sustainable-fsa.com/fsa-payment-files/>

Machine-readable metadata are in [`CITATION.cff`](CITATION.cff); GitHub's **Cite this repository** button (top right of the repo page) renders it as APA or BibTeX.

**Acknowledgment**: This work is part of the [*Enhancing Sustainable Disaster Relief in FSA Programs*](https://www.ars.usda.gov/research/project/?accnNo=444612) project, supported by the USDA Office of the Chief Economist, Office of Energy and Environmental Policy, and the USDA Climate Hubs.

## 📄 License

- **Raw FSA Farm Payment Files data** (USDA): Public Domain (17 USC § 105)
- **Processed data & scripts**: © R. Kyle Bocinsky, released under [CC0](https://creativecommons.org/publicdomain/zero/1.0/) and [MIT License](./LICENSE) as applicable

---

## ⚠️ Disclaimer

This dataset is archived for research and educational use only. It may not reflect current FSA payments and policy. Always consult your **local FSA office** for the latest program guidance.

To locate your nearest USDA Farm Service Agency office, use the USDA Service Center Locator:

🔗 [**USDA Service Center Locator**](https://offices.sc.egov.usda.gov/locator/app)

---

## 👏 Acknowledgment

This project is part of:

**[*Enhancing Sustainable Disaster Relief in FSA Programs*](https://www.ars.usda.gov/research/project/?accnNo=444612)**  
Supported by USDA OCE/OEEP and USDA Climate Hubs  
Prepared by the [Montana Climate Office](https://climate.umt.edu)

---

## 📬 Contact

**R. Kyle Bocinsky**  
Director of Climate Extension  
Montana Climate Office  
📧 [kyle.bocinsky@umontana.edu](mailto:kyle.bocinsky@umontana.edu)  
🌐 [https://climate.umt.edu](https://climate.umt.edu)
